Approximating the matrix exponential derivative of the CTMC generator matrix, with application to global SARS-CoV-2

### 2023-03-16 manuscript/main.tex
stopping 4 now

 Let data $\y$ be the $N$-vector with elements $y_n\in \{1,\dots,S\}$, the observed states of each observed biological specimen $n \in \{1,\dots,N\}$.  At any node $v$, the likelihood takes the form \citep{ji2020gradients}
 \begin{align}
p(\y|\ttheta) = \p_v^T \q_v \, ,
 \end{align}
where $\p_v$ and $\q_v$ are the post-order and pre-order partial likelihood vectors, respectively.  The former describes the probability of the observed states for all observed specimens (i.e., leaf nodes) that descend from node $v$, conditioned on the state at node $v$. The latter describes analogous probabilities for all observed specimens not descending from node $v$.  If we suppose that node $u$ gives rise to two child nodes, $v$ and $w$, then 
\begin{align}\label{eq:recursions}
	\p_u = \P_v \p_v \circ \P_w \p_w\, ,  \quad \q_v= \P_v^T \left( \q_u \circ  \P_w \p_w\right) \, .
\end{align}
Using the chain rule and the fact that $\p_v$ does not depend on $\P_v$, one may write the likelihood's derivative with respect to a single parameter $\theta_d$ thus:
\begin{align}\label{eq:phyloGrad}
	\frac{\partial}{\partial \theta_d}  p(\y|\ttheta) 
	&=   \sum_{v=1}^{2N-2} \mbox{tr} \left( \frac{\partial (\p_v^T\q_v)}{\partial \P_v} \frac{\partial \P^T_v}{\partial \theta_d} \right)
	= \sum_{v=1}^{2N-2} \mbox{tr} \left(  (\q_u\circ \P_w\p_w) \p_v^T \frac{\partial e^{\gamma t_v \QQ^T}}{\partial \theta_d}\right)  \\ \nonumber
	&=\sum_{v=1}^{2N-2} \p_v^T \left(\sum_{s=1}^S \sum_{s'=1}^S \frac{\partial e^{\gamma t_v \QQ^T}}{\partial q_{ss'}} \frac{\partial q_{ss'}}{\partial \theta_d} \right) (\q_u\circ \P_w\p_w)  \, .
\end{align}
Whereas the recursions of \eqref{eq:recursions} facilitate fast likelihood computation, inferring $\ttheta$ using gradient-based approaches such as HMC (Section \ref{sec:surrHmc}) requires a large number of repeated evaluations of the matrix exponential derivative.  These computations become particularly onerous when one opts for a gold-standard mixed-effect model \textcolor{red}{(MAGEE)} and specifies
\begin{align}\label{eq:mixed}
\log q_{ss'} = b_{ss'} + \epsilon_{ss'} \,, \quad s\neq s'\, .
\end{align} 
Here, the fixed-effects $b_{ss'}$ are elements of some non-random matrix $\B$, and the random-effects $\epsilon_{ss'}$ are mutually independent \emph{a priori} and inferred as model parameters.  The dimension $D$ of $\ttheta$ in this model is $\mathcal{O}(S^2)$, so the usual $\mathcal{O}(DNS^5)$ cost of the matrix expoential derivative becomes a massive $\mathcal{O}(NS^7)$.  In this context, \textcolor{red}{MAGEE} show that the first-approximation \eqref{eq:firstOrder} achieves a considerable speedup over the exact derivative, while facilitating high-quality proposals in the context of surrogate HMC.  In the following section, we use this method to analyze the global spread of SARS-CoV-2 and show that the first-order approximation maintains its performance for an even higher-dimensional problem than previously considered.

@@

>>>>

 Let data $\y$ be the $N$-vector with elements $y_n\in \{1,\dots,S\}$, the observed states of each observed biological specimen $n \in \{1,\dots,N\}$.  At any node $v$, the likelihood takes the form \citep{ji2020gradients}
 \begin{align}
p(\y|\ttheta) = \p_v^T \q_v \, ,
 \end{align}
where $\p_v$ and $\q_v$ are the post-order and pre-order partial likelihood vectors, respectively.  The former describes the probability of the observed states for all observed specimens (i.e., leaf nodes) that descend from node $v$, conditioned on the state at node $v$. The latter describes analogous probabilities for all observed specimens not descending from node $v$.  If we suppose that node $u$ gives rise to two child nodes, $v$ and $w$, then 
\begin{align}\label{eq:recursions}
	\p_u = \P_v \p_v \circ \P_w \p_w\, ,  \quad \q_v= \P_v^T \left( \q_u \circ  \P_w \p_w\right) \, .
\end{align}
Using the chain rule and the fact that $\p_v$ does not depend on $\P_v$, one may write the likelihood's derivative with respect to a single parameter $\theta_d$ thus:
\begin{align}\label{eq:phyloGrad}
	\frac{\partial}{\partial \theta_d}  p(\y|\ttheta) 
	&=   \sum_{v=1}^{2N-2} \mbox{tr} \left( \frac{\partial (\p_v^T\q_v)}{\partial \P_v} \frac{\partial \P^T_v}{\partial \theta_d} \right)
	= \sum_{v=1}^{2N-2} \mbox{tr} \left(  (\q_u\circ \P_w\p_w) \p_v^T\left( \frac{\partial e^{\gamma t_v \QQ}}{\partial \theta_d}\right)^T\right)  \\ \nonumber
	&=\sum_{v=1}^{2N-2} \p_v^T \left(\sum_{s=1}^S \sum_{s'=1}^S \frac{\partial e^{\gamma t_v \QQ}}{\partial q_{ss'}} \frac{\partial q_{ss'}}{\partial \theta_d} \right)^T (\q_u\circ \P_w\p_w)  \, .
\end{align}
Whereas the recursions of \eqref{eq:recursions} facilitate fast likelihood computation, inferring $\ttheta$ using gradient-based approaches such as HMC (Section \ref{sec:surrHmc}) requires a large number of repeated evaluations of the matrix exponential derivative.  These computations become particularly onerous when one opts for a gold-standard mixed-effect model \textcolor{red}{(MAGEE)} and specifies
\begin{align}\label{eq:mixed}
\log q_{ss'} = b_{ss'} + \epsilon_{ss'} \,, \quad s\neq s'\, .
\end{align} 
Here, the fixed-effects $b_{ss'}$ are elements of some non-random matrix $\B$, and the random-effects $\epsilon_{ss'}$ are mutually independent \emph{a priori} and inferred as model parameters.  The dimension $D$ of $\ttheta$ in this model is $\mathcal{O}(S^2)$, so the usual $\mathcal{O}(DNS^5)$ cost of the matrix expoential derivative becomes a massive $\mathcal{O}(NS^7)$.  In this context, \textcolor{red}{MAGEE} show that the first-approximation \eqref{eq:firstOrder} achieves a considerable speedup over the exact derivative, while facilitating high-quality proposals in the context of surrogate HMC.  In the following section, we use this method to analyze the global spread of SARS-CoV-2 and show that the first-order approximation maintains its performance for an even higher-dimensional problem than previously considered.

@@

In the context of a Bayesian analysis, we specify independent normal priors on $\theta_1$, $\theta_2$ and $\theta_3$ with means of 0 and variances of 2.  We also assume that the 1,892 random-effects $\epsilon_{ss'}$ follow sparsity inducing Bayesian bridge priors with global scale parameter $\tau$ and exponent $\alpha=0.25$.  Here, we follow \citet{nishimura2022shrinkage} and specify a Gamma prior on $\tau^{-\alpha}$ with a shape parameter of 1 and a rate parameter of 2. Finally, we place a flat prior on the rate scalar $\gamma$.  Inferring the posterior distribution of all 1,897 model parameters requires an advanced MCMC strategy. Namely, we adopt an HMC-within-Gibbs approach, updating the scalars $\gamma$ and $\tau$ independently but updating all 1,895 regression parameters (both fixed- and random-effects) using surrogate HMC accomplished with the first-order approximation
\begin{align*}
  \gamma t_v  e^{\gamma t_v \QQ} \JJ_{ss'}    \approx \frac{\partial e^{\gamma t_v \QQ}}{\partial q_{ss'}} 
\end{align*}
within \eqref{eq:phyloGrad}.

We generate 10 million MCMC samples in this manner, saving 1 in 10,000 Markov chain states in order to conserve memory.  
\begin{itemize}
	\item ESS
	\item Figure 2
	\item Tree
\end{itemize}

In addition to these scientific questions of interest, we are interested in the performance of the first-order approximation as a surrogate gradient for HMC in such a high-dimensional setting. To do isolate this performance, we temporarily fix the Bayesian bridge global-scale $\tau$ at 2.5$\times10^{-5}$.  We generate
